\LARGE \bf Planning with Attitude

### 2020-10-11 paper/main.tex
edits + comments

@@

\begin{abstract}
Planning trajectories for floating-base robotic systems that experience
large attitude changes is challenging due to the nontrivial group structure of 3D
rotations. This paper introduces a powerful and accessible approach for 
optimization-based planning on the space of rotations based on vector calculus and
linear algebra. We demonstrate the effectiveness of the approach by adapting Newton's
method to solve the canonical Wahba's problem, and modifying the ALTRO solver for 
constrained nonlinear trajectory optimization to plan directly on the space of unit 
quaternions, achieving superior convergence on problems involving significant changes 
in attitude.
\keywords{motion planning and control, quaternions, optimal control, 
linear quadratic regulator
}
\end{abstract}

@@

    Many useful robotic systems---including quadrotors, airplanes, satellites, autonomous
    underwater vehicles, and quadrupeds---can perform arbitrarily large three-dimensional
    translations and rotations as part of their normal operation. While representing
    translations is straightforward and intuitive, effectively representing the
    nontrivial group structure of 3D rotations has been a topic of study for many
    decades. Although we can intuitively deduce that rotations are three-dimensional, a
    globally non-singular three-parameter representation of the space of rotations does
    not exist \cite{stuelpnagel1964parametrization}. As a result, when parameterizing
    rotations, we must either a) pick a three-parameter representation and deal with
    discontinuities, or b) pick a higher-dimensional representation and deal with
    constraints between the parameters. While simply representing attitude is nontrivial,
    generating and tracking motion plans for floating-base systems is an even more
    challenging problem.

@@

    More recently, these ideas have been extended to trajectory generation
    \cite{Zefran1998}, sample-based motion planning \cite{Zefran1999,Kuffner2004}, and
    optimal control. Approaches to optimal control on attitude problems include
    analytical methods applied to satellites \cite{Spindler1998}, discrete mechanics
    \cite{Kobilarov2011,Kobilarov2014, Lee2008}, a combination of sampling-based planning
    and constrained trajectory optimization for satellite formations \cite{Garcia2005,
    Aoude2008}, projection operators \cite{Saccon2013}, or more general theory for
    optimization on manifolds \cites{watterson2018trajectory}. Nearly all of these
    methods rely heavily on principles from differential geometry and Lie group theory;
    however, despite these works, many recent papers in the robotics community continue
    to apply traditional methods for motion planning and control with no regard for the
    group structure of rigid body motion \cite{Alothman2016,deCrousaz2015,
    Williams2017,Geisert2016}.

    In this paper, we make a departure from previous approaches to geometric planning and 
    control that rely heavily on ideas and notation from differential geometry, 
    and instead use only basic mathematical tools from linear algebra and calculus that 
    should be familiar to most roboticists. 
    Similar to \cite{Mandic2011,Xu2016}, in Sec. \ref{sec:Quaternion_Calculus} we introduce 
    a quaternion differential calculus, but take a significantly simpler and more general 
    approach, enabling straight-forward adaptation of 
    existing algorithms to systems with quaternion states. 
    To make this concrete, in Sec. \ref{sec:Wahbas} we apply this method to the canonical
    Wahba's problem and demonstrate dramatically superior convergence to approaches that
    fail to properly account for the group structure. 
    In Sec. \ref{sec:trajopt} we extend these ideas to the problem of trajectory optimization,
    and detail modifications to ALTRO, a state-of-the-art constrained trajectory optimization
    solver, and demonstrate the performance gains on several constrained benchmark problems.
    To our knowledge, there does not currently exist a solver that is capable of leveraging
    the unique Markovian structure of the fixed-horizon trajectory optimization problem while
    correctly accounting for the group structure of 3D rotations.
    In summary, our contributions include:

    \begin{itemize}
        \item A unified approach to quaternion differential calculus entirely based on
        standard vector calculus and linear algebraic operations, and
        \item a fast and efficient solver for trajectory optimization problems with 
        attitude dynamics and nonlinear constraints that correctly accounts for the group
        structure of 3D rotations during the solve.
    \end{itemize}

@@

    \subsection{Unit Quaternions} \label{sec:quaternions}
        We leverage the fact that quaternions are linear operators and that the space of 
        quaternions $\mathbb{H}$ is isomorphic to $\R^4$ to explicitly 
        represent---following the Hamilton convention---a quaternion $\q \in \mathbb{H}$ as 
        a standard vector $q \in \R^4 := [q_s \;\; q_v^T]^T$ where $q_s \in \R$ and 
        $q_v \in \R^3$ are the scalar and vector part of the quaternion, respectively.

@@

    \subsection{Rigid Body Dynamics} \label{sec:rigidbody_dynamics}
        In the current work we will restrict our focus to rigid bodies moving freely in 3D 
        space. That is, we consider systems with dynamics of the following form:
        \begin{equation} \label{eq:rigid_body_dynamics}
            x = \begin{bmatrix} r \\ R \\ v \\ \omega \end{bmatrix}, \quad 
            \dot{x} = \begin{bmatrix} 
                v \\ 
                \half \q \otimes \hat{\omega} = \half L(q) H \omega \\ 
                \frac{1}{m} F_G(x,u) \\ 
                J^{-1}(\tau_L(x,u) - \omega \times J \omega) 
            \end{bmatrix}
        \end{equation}
        where $x$ and $u$ are the state and control vectors, $r \in \R^3$ is the position, 
        $R \in SO(3)$ is the attitude, $v \in \R^3$ is the linear velocity, and 
        $\omega \in \R^3$ is the angular velocity. $m \in \R$ is the mass, 
        $J \in \R^{3\times3}$ is the inertia matrix, $F_G(x,u) \in \R^3$ are the forces in the 
        global frame, and $\tau_L(x,u)$ are the moments in the local (body) frame.

@@

        The fact that differential unit quaternions are three-dimensional should make
        intuitive sense: Rotations are inherently three-dimensional and differential
        rotations should live in the same space as angular velocity, i.e. $\R^3$.

@@

        Let $\phi \in \R^3$ be a differential rotation applied to a function with
        quaternion inputs $y = h(q): \Q \to \R^p$, such that
        \begin{equation} \label{eq:vector_function}
            y + \delta y = h(L(q) \varphi(\phi)) \approx h(q) +  \nabla h(q) \phi.
        \end{equation}
        We can calculate the Jacobian $\nabla h(q) \in \R^{p \times 3}$ by
        differentiating \eqref{eq:vector_function} with respect to $\phi$, evaluated at
        $\phi = 0$:
        \begin{equation} \label{eq:quat_gradient}
            \nabla h(q) = \pdv{h}{q} L(q) H := \pdv{h}{q} G(q) 
                        = \pdv{h}{q} \begin{bmatrix} 
                            -q_v^T \\ 
                            sI_3 + \skewmat{q_v}
                        \end{bmatrix}
        \end{equation}
        where $G(q) \in \R^{4 \times 3}$ is the \textit{attitude Jacobian}, which
        essentially becomes a ``conversion factor'' allowing us to apply results from
        standard vector calculus to the space of unit quaternions. This form is
        particularly useful in practice since $\pdv*{h}{q} \in \R^{p \times 4}$ can be
        obtained using finite difference or automatic differentiation.
        As an aside, although we have used Rodrigues parameters, $G(q)$ is actually the
        same (up to a constant scaling factor) for any choice of three-parameter attitude
        representation.

@@

	    \begin{equation} \label{eq:quat_hessian}
            \nabla^2 h(q) = G(q)^T \pdv[2]{h}{q} G(q) + I_3 \pdv{h}{q}q,
	    \end{equation}
	    where the second term comes from the second derivative of $\varphi(\phi)$.
	    Similar to $G(q)$, this ends up being the same (up to a scaling factor) for any
        choice of three-parameter attitude representation.

    \subsection{Jacobian of Quaternion-Valued Functions}
        We now consider the case of a function that maps unit quaternions to unit
        quaternions, $q' = f(q) : \Q \to \Q$. Here we must also consider the non-trivial
        effect of a differential value applied to the output, i.e.:
        \begin{equation} \label{eq:dqoutput}
            L(q') \varphi(\phi') = f(L(q)\varphi(\phi)) .
        \end{equation}
        Solving \eqref{eq:dqoutput} for $\phi'$ we find,
        \begin{equation} \label{eq:phiprime}
            \phi' = \varphi^{-1} \left( L(q')^T f(L(q)\varphi(\phi)) \right) \approx \nabla f(q) \, \phi.
        \end{equation}
        Finally, the desired Jacobian is obtained by taking the derivative of
        \eqref{eq:phiprime} with respect to $\phi$:
        \begin{equation} \label{eq:quat_jacobian}
            \nabla f(q) = H^T L(q')^T \pdv{f}{q} L(q) H = G(q')^T \pdv{f}{q} G(q).
        \end{equation}
        The leading $G(q')^T$ comes from the fact that as $\phi' \to 0$, $L(q') f(q) \to
        I_q$, where $I_q$ is the quaternion identity. Differentiating through the inverse
        map, evaluated at the quaternion identity, we find that $\pdv*{\varphi^{-1}}{q}
        \to H^T$ for any three-parameter attitude representation.

@@

    Newton's method uses derivative information about a function to iteratively
    approximate its roots. For unconstrained systems, this method is highly effective,
    and can exhibit quadratic convergence. For constrained systems, the updated parameter
    can be projected back onto the feasible set at each iteration, but without the same
    convergence guarantees. For the constraints on $SO(3)$, Newton's method struggles to
    converge past a certain threshold due to this projection. By leveraging the
    quaternion calculus introduced, Newton's method can be modified to implicitly account
    for these constraints. To demonstrate this, we will examine Wahba's Problem. In 1965,
    Grace Wahba proposed the criterion for a least squares estimate of a spacecraft's
    attitude from vector measurements \cite{markley2014fundamentals}. We will solve this
    problem using a standard nonlinear least squares method, as well as a method that
    exploits the true group structure of $SO(3)$ using the quaternion calculus presented
    here.

    \subsection{Methodology}

    Given known vectors in some inertial frame, ${}^N v_i$, and measurements of these
    vectors in some body fixed frame, ${}^B v_i$, our goal is to determine the relative
    rotation from the body to inertial frame ${}^N q^B$, expressed as a quaternion. We
    can define Wahba's loss function as the following: 
    \begin{align} \label{eq:wahba_loss}
        L &= \sum_i w_i \norm{{}^N v_i - {}^N A(q)^B \,\, ^Bv_i }_2^2. = \norm{r_i(q)}_2^2 
    \end{align} 
    where $r_i(q)$ is the residual vector.

    We can solve for ${}^N q^B$ using a nonlinear least squares method minimizing
    Wahba's loss function:
    \begin{mini*}
        {q}{ \norm{r(q)}_2^2 }{}{}
        \addConstraint { q\in SO(3).}
    \end{mini*}

    Following the typical approach for Newton's method, we minimize \eqref{eq:wahba_loss}
    by setting the gradient to zero:
    \begin{equation} \label{eq:newton_foc}
        \begin{aligned}
            \pdv{L}{q}^T &= \sum_i \pdv{r(q)}{q}^T r_i{q} := J^T r(q) = 0 \\
              &= \sum_i (-2 H^T R(q)^T R({}^B \hat{v}_i)^T r(q).
        \end{aligned}
    \end{equation}
    which can be obtained from the chain rule and \eqref{eq:quaternion_rotation}.

@@

    \subsection{Results}
    As illustrated in Figure \ref{fig:wahba_convergence}, it is clear that by projecting 
    back onto the unit quaternion at each iteration make initial progress, but fails to 
    exhibit the quadratic convergence typical of a Newton method. By optimizing directly 
    in the space of differential quaternions, we achieve the expected quadratic convergence.
    It should also be clear from the previous section that the adaptations to the original
    Newton method are simple and straightforward, highlighting both the effectiveness and
    value of the proposed approach.
    \begin{figure}

@@

\section{Trajectory Optimization on $\R^n \times SO(3)$} \label{sec:trajopt}
    Here we outline the modifications to the ALTRO solver \cite{howell2019altro}, to
    solve trajectory optimization problems for rigid bodies, which extends easily to
    arbitrary systems in $\R^n \times SO(3)$. ALTRO is an efficient solver for constrained
    nonlinear optimization problems that uses iterative LQR (iLQR) with an augmented
    Lagrangian framework.

    We consider a trajectory optimization problem of the form:
    \begin{mini}[2]
        {x_{0:N},u_{0:N-1}}{\ell_f(x_N) + \sum_{k=0}^{N-1} \ell(x_k,u_k) }{}{}
        \addConstraint{x_{k+1} = f(x_k,u_k)}
        \addConstraint{g_k(x_k,u_k)}{\leq 0}
        \addConstraint{h_k(x_k,u_k)}{=0}
        \label{discrete_trajopt},
    \end{mini}
    where $x$ and $u$ are the state and control vectors as described in Sec. \ref{sec:rigidbody_dynamics},
    $f$ are the dynamics as defined in \eqref{eq:rigid_body_dynamics}, $\ell$ is a general
    nonlinear cost function,
    $N$ is the number of time steps, and $g_k$, $h_k$ are general nonlinear inequality and 
    equality constraints.

@@

    We begin by linearizing the dynamics about the reference trajectory using
    \eqref{eq:quat_jacobian}. Our linearized error dynamics become
    \begin{equation} \label{eq:linearized_dynamics}
        \delta x_{k+1} = A_k \delta x_k + B_k \delta u_k 
    \end{equation}
    where \begin{equation}
        \begin{aligned}
            A_k = E(\bar{x}_{k+1})^T \pdv{f}{x}|_{\bar{x}_k,\bar{u}_k} E(\bar{x}_k), \\
            B_k = E(\bar{x}_{k+1})^T \pdv{f}{u}|_{\bar{x}_k,\bar{u}_k},
        \end{aligned}
    \end{equation}
    and $\delta x_k \in \R^{12}$ and $E(x_k) \in \R^{12 \times 13}$ are the state error and state error Jacobian:
    \begin{equation} \label{eq:state_error}
        \setlength\arraycolsep{1pt}
        \delta x_k = \begin{bmatrix} 
            r_k - \bar{r}_k \\ \varphi^{-1}(\bar{\q}_k^{-1} \otimes \q_k) \\ v_k - \bar{v}_k \\ \omega_k - \bar{\omega}_k 
        \end{bmatrix}\!, \;
        E(x) = \begin{bmatrix}
            I_3 & & & \\
            & G(q) & & \\
            & & I_3 & \\
            & & & I_3 \\
        \end{bmatrix}\!.
    \end{equation}

    By applying \eqref{eq:quat_gradient} and \eqref{eq:quat_hessian} to our nonlinear cost
    functions $\ell$ and 
    \eqref{eq:quat_jacobian} to the nonlinear constraint functions $g_k$ and $h_k$, we can calculate
    the second-order expansion 
    \begin{multline}
        \delta \ell(x,u) \approx  \half \delta x^T \ell_xx \delta x
            + \half \delta u^T \ell_{uu} \delta u + \delta_u^T \ell_{ux} \delta u \\
            + \ell_x^T \delta x^T + \ell_u^T \delta u 
    \end{multline}
    of the augmented Lagrangian cost function:
    \begin{equation}
        \mathcal{L}_A = \mathcal{L}_N(x_N,\lambda_N,\mu_N) + 
            \sum_{k=0}^{N-1} \mathcal{L}_k(x_k,u_k,\lambda_k,\mu_k)
    \end{equation}
    where
    \begin{equation}
        \mathcal{L}_k(x,u,\lambda,\mu) = \ell(x,u) + 
            (\lambda + \half I_\mu c(x,u))^T c(x,u),
    \end{equation}
    with $c(x,u)$ being the concatenation of the constraints $f,g$, and $h$ at a given time step,
    and $I_\mu$ the penalty matrix.

@@

    \subsection{Quaternion Cost Functions} \label{sec:cost_functions}
        In addition to the straight-forward modifications to the ALTRO algorithm itself, we
        need to carefully consider the types of cost functions we minimize. We frequently
        minimize costs that penalize distance from a goal state, e.g. $\half (x-x_g)^T Q
        (x-x_g)$; however, na\"ive substraction of unit quaternions is ill-defined. We
        propose two different cost functions that accomplish similar behavior.
        For sake of clarity and space, we only consider the costs on the quaternion
        variables: costs on the other states and the control variables remain unaffected.

@@

        \subsubsection{Geodesic Distance} \label{sec:geodesic}
            Alternatively, we can use the geodesic distance between two quaternions
            \cite{Kuffner2004},
            \begin{equation} \label{eq:quat_geodesic}
                J_\text{geo} = (1-\abs{q_g^T q}) ,
            \end{equation}
            whose gradient and Hessian are,
            \begin{align}
                \nabla J_\text{geo} &= \pm q_g^T G(q) \\
                \nabla^2 J_\text{geo} &= \pm I_3 q_g^T q ,
            \end{align}
            where the sign of the Hessian corresponds to the sign of $q_d^T q$.

>>>>

\DeclareMathOperator{\sign}{sign}

@@

\begin{abstract}
Planning trajectories for floating-base robotic systems that experience
large attitude changes is challenging due to the nontrivial group structure of 3D
rotations. This paper introduces a powerful and accessible approach for 
optimization-based planning on the space of rotations using only standard linear algebra and vector calculus. We demonstrate the effectiveness of the approach by adapting Newton's
method to solve the canonical Wahba's problem, and modifying the trajectory optimization solver ALTRO to plan directly on the space of unit 
quaternions, achieving superior convergence on problems involving significant changes 
in attitude.
%\keywords{trajectory optimization, motion planning, control, quaternions}

\end{abstract}

@@

    Many robotic systems---including quadrotors, airplanes, satellites, autonomous
    underwater vehicles, and quadrupeds---can perform arbitrarily large three-dimensional
    translations and rotations as part of their normal operation. While representing
    translations is straightforward and intuitive, effectively representing the
    nontrivial group structure of 3D rotations has been a topic of study for many
    decades. Although we can intuitively deduce that rotations are three-dimensional, a
    globally non-singular three-parameter representation of the space of rotations does
    not exist \cite{stuelpnagel1964parametrization}. As a result, when parameterizing
    rotations, we must either a) choose a three-parameter representation and deal with
    singularities and discontinuities, or b) choose a higher-dimensional representation and deal with
    constraints between the parameters. While simply representing attitude is nontrivial,
    generating and tracking motion plans for floating-base systems is an even more
    challenging problem.

@@

    More recently, these ideas have been extended to trajectory generation
    \cite{Zefran1998}, sample-based motion planning \cite{Zefran1999,Kuffner2004}, and
    optimal control. Approaches to optimal control with attitude states include
    analytical methods applied to satellites \cite{Spindler1998}, discrete mechanics
    \cite{Kobilarov2011,Kobilarov2014, Lee2008}, a combination of sampling-based planning
    and constrained trajectory optimization for satellite formations \cite{Garcia2005,
    Aoude2008}, projection operators \cite{Saccon2013}, or more general theory for
    optimization on manifolds \cites{watterson2018trajectory}. Nearly all of these
    methods rely heavily on principles from differential geometry and Lie group theory;
    however, despite these works, many recent papers in the robotics community continue
    to naively apply standard methods for motion planning and control with no regard for the
    group structure of rigid body motion \cite{Alothman2016,deCrousaz2015,
    Williams2017,Geisert2016}.

    In this paper, we make a departure from previous approaches to geometric planning and 
    control that rely heavily on ideas and notation from differential geometry, 
    and instead use only basic mathematical tools from linear algebra and vector calculus that 
    should be familiar to most roboticists. In Sec. \ref{sec:Quaternion_Calculus} we introduce 
    an approach to quaternion differential calculus similar to \cite{Mandic2011,Xu2016}, but significantly simpler and more general, enabling straight-forward adaptation of 
    existing algorithms to systems with quaternion states. 
    For concreteness, we then apply our method to the canonical
    Wahba's problem~\cite{Wahba1965} in Sec. \ref{sec:Wahbas}, and demonstrate superior convergence to approaches that
    fail to properly account for the group structure. 
    In Sec. \ref{sec:trajopt} we extend these ideas to the problem of trajectory optimization,
    and detail modifications to ALTRO, a state-of-the-art constrained trajectory optimization
    solver, and demonstrate performance gains on several benchmark problems.
    %To our knowledge, there does not currently exist a solver that is capable of leveraging the unique Markovian structure of the fixed-horizon trajectory optimization problem while correctly accounting for the group structure of 3D rotations.
    In summary, our contributions include:

    \begin{itemize}
        \item A unified approach to quaternion differential calculus entirely based on standard linear algebra and vector calculus.
        \item Derivation of a Newton-based algorithm for nonlinear optimization directly on the space of unit quaternions using our notation.
        \item a fast and efficient solver for trajectory optimization problems with 
        attitude dynamics and nonlinear constraints that correctly accounts for the group
        structure of 3D rotations.
    \end{itemize}

@@

    \subsection{Unit Quaternions} \label{sec:quaternions}
        We leverage the fact that quaternions are linear operators and that the space of 
        quaternions $\mathbb{H}$ is isomorphic to $\R^4$ to explicitly 
        represent---following the Hamilton convention---a quaternion $\q \in \mathbb{H}$ as 
        a standard vector $q \in \R^4 := [q_s \;\; q_v^T]^T$ where $q_s \in \R$ and 
        $q_v \in \R^3$ are referred to as the scalar and vector parts of the quaternion, respectively.

@@

    \subsection{Rigid Body Dynamics} \label{sec:rigidbody_dynamics}
        For clarity, we will restrict our attention to rigid bodies moving freely in 3D 
        space. That is, we consider systems with dynamics of the following form:
        \begin{equation} \label{eq:rigid_body_dynamics}
            x = \begin{bmatrix} r \\ R \\ v \\ \omega \end{bmatrix}, \quad 
            \dot{x} = \begin{bmatrix} 
                v \\ 
                \half \q \otimes \hat{\omega} = \half L(q) H \omega \\ 
                \frac{1}{m} F_G(x,u) \\ 
                J^{-1}(\tau_L(x,u) - \omega \times J \omega) 
            \end{bmatrix}
        \end{equation}
        where $x$ and $u$ are the state and control vectors, $r \in \R^3$ is the position, 
        $R \in SO(3)$ is the attitude \todo{why didn't we just make this $q$?}, $v \in \R^3$ is the linear velocity, and 
        $\omega \in \R^3$ is the angular velocity. $m \in \R$ is the mass, 
        $J \in \R^{3\times3}$ is the inertia matrix, $F_G(x,u) \in \R^3$ are the forces in the 
        global frame, and $\tau_L(x,u)$ are the moments in the local (body) frame \todo{Can we stick with just the term ``body'' throughout the paper (and maybe change $L$ to $B$ here)?}.

@@

        The fact that differential unit quaternions are three-dimensional should make
        intuitive sense: Rotations are inherently three-dimensional and differential
        rotations should live in the same space as angular velocities, i.e. $\R^3$.

@@

        Let $\phi \in \R^3$ be a differential rotation applied to a function with
        quaternion inputs $y = h(q): \Q \to \R^p$, such that
        \begin{equation} \label{eq:vector_function}
            y + \delta y = h(L(q) \varphi(\phi)) \approx h(q) +  \nabla h(q) \phi.
        \end{equation}
        We can calculate the Jacobian $\nabla h(q) \in \R^{p \times 3}$ by
        differentiating \eqref{eq:vector_function} with respect to $\phi$, evaluated at
        $\phi = 0$:
        \begin{equation} \label{eq:quat_gradient}
            \nabla h(q) = \pdv{h}{q} L(q) H := \pdv{h}{q} G(q) 
                        = \pdv{h}{q} \begin{bmatrix} 
                            -q_v^T \\ 
                            sI_3 + \skewmat{q_v}
                        \end{bmatrix}
        \end{equation}
        where $G(q) \in \R^{4 \times 3}$ is the \textit{attitude Jacobian}, which
        essentially becomes a ``conversion factor'' allowing us to apply results from
        standard vector calculus to the space of unit quaternions. This form is
        particularly useful in practice since $\pdv*{h}{q} \in \R^{p \times 4}$ can be
        obtained using finite differences or automatic differentiation.
        As an aside, although we have used Rodrigues parameters, $G(q)$ is actually the
        same (up to a constant scalar factor) for any choice of three-parameter attitude
        representation.

@@

	    \begin{equation} \label{eq:quat_hessian}
            \nabla^2 h(q) = G(q)^T \pdv[2]{h}{q} G(q) + I_3 \pdv{h}{q}q,
	    \end{equation}
	    where the second term comes from the second derivative of $\varphi(\phi)$.
	    Similar to $G(q)$, this expression is the same (up to a constant scalar factor) for any
        choice of three-parameter attitude representation.

    \subsection{Jacobian of Quaternion-Valued Functions}
        We now consider the case of a function that maps unit quaternions to unit
        quaternions, $q' = f(q) : \Q \to \Q$. \todo{Let's make sure we're bing consistent with our usage of $\Q$ vs. $\mathbb{H}$.} Here we must also consider the non-trivial
        effect of a differential rotation applied to the output, i.e.:
        \begin{equation} \label{eq:dqoutput}
            L(q') \varphi(\phi') = f(L(q)\varphi(\phi)) .
        \end{equation}
        Solving \eqref{eq:dqoutput} for $\phi'$ we find,
        \begin{equation} \label{eq:phiprime}
            \phi' = \varphi^{-1} \left( L(q')^T f(L(q)\varphi(\phi)) \right) \approx \nabla f(q) \, \phi.
        \end{equation}
        Finally, the desired Jacobian is obtained by taking the derivative of
        \eqref{eq:phiprime} with respect to $\phi$:
        \begin{equation} \label{eq:quat_jacobian}
            \nabla f(q) = H^T L(q')^T \pdv{f}{q} L(q) H = G(q')^T \pdv{f}{q} G(q).
        \end{equation}
        The leading $G(q')^T$ comes from the fact that as $\phi' \to 0$, $L(q') f(q) \to
        I_q$, where $I_q$ is the quaternion identity. Once again, \eqref{eq:quat_jacobian} holds (up to a constant) for any three-parameter attitude representation.

        %Differentiating through the inverse map, evaluated at the quaternion identity, we find that $\pdv*{\varphi^{-1}}{q}\to H^T$ for any three-parameter attitude representation.

@@

\todo{I would re-work this section to only talk about the details of doing it the ``correct'' way (i.e. consistent with the previous section). Then just call the standard version a ``naive Newton method in which the quaternion is re-normalized at each step'' and don't bother writing any of the mathematical details for that version. I think it will get confusing if you mix derivative types/notation.}

    Newton's method uses derivative information about a function to iteratively
    approximate its roots. For unconstrained systems, this method is highly effective,
    and can exhibit quadratic convergence. For constrained systems, the updated parameter
    can be projected back onto the feasible set at each iteration, but without the same
    convergence guarantees. For the constraints on $SO(3)$, Newton's method struggles to
    converge past a certain threshold due to this projection. By leveraging the
    quaternion calculus introduced, Newton's method can be modified to implicitly account
    for these constraints. To demonstrate this, we will examine Wahba's Problem. In 1965,
    Grace Wahba proposed the criterion for a least squares estimate of a spacecraft's
    attitude from vector measurements \cite{Wahba1965, markley2014fundamentals}. We will solve this
    problem using a standard nonlinear least squares method, as well as a method that
    exploits the true group structure of $SO(3)$ using the quaternion calculus presented
    here.

    \subsection{Methodology}
\todo{Let's keep our notation and language consistent for body vs. inertial frames. I propose ``body'' (B) and ``inertial'' or ``world'' (W), which is the standard in robotics.}
    Given known vectors in some inertial frame, ${}^N v_i$, and measurements of these
    vectors in some body fixed frame, ${}^B v_i$, our goal is to determine the relative
    rotation from the body to inertial frame ${}^N q^B$, expressed as a quaternion. We
    can define Wahba's loss function as the following: 
    \begin{align} \label{eq:wahba_loss}
        L &= \sum_i w_i \norm{{}^N v_i - {}^N A(q)^B \,\, ^Bv_i }_2^2. = \norm{r_i(q)}_2^2 
    \end{align} 
    \todo{Let's leave the $w_i$ off for clarity since you're not using it later.}
    where $r_i(q)$ is the residual vector.

    We can solve for ${}^N q^B$ using a nonlinear least squares method minimizing
    Wahba's loss function:
    \begin{mini*}
        {q}{ \norm{r(q)}_2^2 }{}{}
        \addConstraint { q\in SO(3).}
    \end{mini*}
    \todo{This should be $q \in \Q$ here, since unit quaternions are not isomorphic to $SO(3)$ (rotation matrices).}
    Following the typical approach for Newton's method, we minimize \eqref{eq:wahba_loss}
    by setting the gradient to zero:
    \begin{equation} \label{eq:newton_foc}
        \begin{aligned}
            \pdv{L}{q}^T &= \sum_i \pdv{r(q)}{q}^T r_i{q} := J^T r(q) = 0 \\
              &= \sum_i (-2 H^T R(q)^T R({}^B \hat{v}_i)^T r(q).
        \end{aligned}
    \end{equation}
    which can be obtained from the chain rule and \eqref{eq:quaternion_rotation}.

@@

    \subsection{Results}
    As illustrated in Figure \ref{fig:wahba_convergence}, it is clear that by projecting 
    back onto the unit quaternion at each iteration make initial progress, but fails to 
    exhibit the quadratic convergence typical of a Newton method. By optimizing directly 
    in the space of unit quaternions, we achieve the expected quadratic convergence.
    It should also be clear from the previous section that the adaptations to the original
    Newton method are simple and straightforward, highlighting both the effectiveness and
    value of the proposed approach.
    \begin{figure}

@@

\section{Trajectory Optimization on $\R^n \times SO(3)$} \label{sec:trajopt}
    Here we outline the modifications to the ALTRO solver \cite{howell2019altro}, to
    solve trajectory optimization problems for rigid bodies, which extends easily to
    arbitrary systems whose state is in $\R^n \times SO(3)$. %ALTRO is an efficient solver for constrained nonlinear optimization problems that uses iterative LQR (iLQR) with an augmented Lagrangian framework.

    We consider trajectory optimization problems of the form,
    \begin{mini}[2]
        {x_{0:N},u_{0:N-1}}{\ell_f(x_N) + \sum_{k=0}^{N-1} \ell(x_k,u_k) }{}{}
        \addConstraint{x_{k+1} = f(x_k,u_k)}
        \addConstraint{g_k(x_k,u_k)}{\leq 0}
        \addConstraint{h_k(x_k,u_k)}{=0}
        \label{discrete_trajopt},
    \end{mini}
    where $x$ and $u$ are the state and control vectors as described in Sec. \ref{sec:rigidbody_dynamics},
    $f$ are the dynamics as defined in \eqref{eq:rigid_body_dynamics}, $\ell$ is a general
    nonlinear cost function,
    $N$ is the number of time steps, and $g_k$, $h_k$ are general nonlinear inequality and 
    equality constraints.

@@

    We begin by linearizing the dynamics about the reference trajectory using
    \eqref{eq:quat_jacobian}. Our linearized error dynamics become
    \begin{equation} \label{eq:linearized_dynamics}
        \delta x_{k+1} = A_k \delta x_k + B_k \delta u_k 
    \end{equation}
    where \begin{equation}
        \begin{aligned}
            A_k = E(\bar{x}_{k+1})^T \pdv{f}{x}|_{\bar{x}_k,\bar{u}_k} E(\bar{x}_k), \\
            B_k = E(\bar{x}_{k+1})^T \pdv{f}{u}|_{\bar{x}_k,\bar{u}_k},
        \end{aligned}
    \end{equation}
    and $\delta x_k \in \R^{12}$ and $E(x_k) \in \R^{12 \times 13}$ are the state error and state-error Jacobian, respectively:
    \begin{equation} \label{eq:state_error}
        \setlength\arraycolsep{1pt}
        \delta x_k = \begin{bmatrix} 
            r_k - \bar{r}_k \\ \varphi^{-1}(\bar{\q}_k^{-1} \otimes \q_k) \\ v_k - \bar{v}_k \\ \omega_k - \bar{\omega}_k 
        \end{bmatrix}\!, \;
        E(x) = \begin{bmatrix}
            I_3 & & & \\
            & G(q) & & \\
            & & I_3 & \\
            & & & I_3 \\
        \end{bmatrix}\!.
    \end{equation}

    By applying \eqref{eq:quat_gradient} and \eqref{eq:quat_hessian} to our nonlinear cost
    functions $\ell$ and 
    \eqref{eq:quat_jacobian} to the nonlinear constraint functions $g_k$ and $h_k$, we can calculate
    the second-order expansion 
    \begin{multline}
        \delta \ell(x,u) \approx  \half \delta x^T \ell_{xx} \delta x
            + \half \delta u^T \ell_{uu} \delta u + \delta_u^T \ell_{ux} \delta u \\
            + \ell_x^T \delta x^T + \ell_u^T \delta u 
    \end{multline}
    of the augmented Lagrangian cost function: \todo{I think you can safely leave out any references to the augmented Lagrangian, etc. and just talk about the cost function.}
    \begin{equation}
        \mathcal{L}_A = \mathcal{L}_N(x_N,\lambda_N,\mu_N) + 
            \sum_{k=0}^{N-1} \mathcal{L}_k(x_k,u_k,\lambda_k,\mu_k)
    \end{equation}
    where
    \begin{equation}
        \mathcal{L}_k(x,u,\lambda,\mu) = \ell(x,u) + 
            (\lambda + \half I_\mu c(x,u))^T c(x,u),
    \end{equation}
    with $c(x,u)$ being the concatenation of the constraints $f,g$, and $h$ at a given time step,
    and $I_\mu$ the penalty matrix.

@@

    \subsection{Quaternion Cost Functions} \label{sec:cost_functions}
        In addition to the straight-forward modifications to the ALTRO algorithm itself, we
        need to carefully consider the types of cost functions we minimize. We frequently
        minimize costs that penalize distance from a goal state, e.g. $\half (x-x_g)^T Q
        (x-x_g)$; However, na\"ive substraction of unit quaternions is ill-defined. We
        propose two different cost functions that accomplish similar behavior.
        For sake of clarity and space, we only consider the costs on the quaternion
        variables: costs on the other states and the control variables remain unaffected.

@@

        \subsubsection{Geodesic Distance} \label{sec:geodesic}
            Alternatively, we can use the geodesic distance between two quaternions:
            \cite{Kuffner2004},
            \begin{equation} \label{eq:quat_geodesic}
                J_\text{geo} = (1-\abs{q_g^T q}) ,
            \end{equation}
            whose gradient and Hessian are,
            \begin{align}
                \nabla J_\text{geo} &= \sign(q_d^T q) q_g^T G(q) \\
                \nabla^2 J_\text{geo} &= \sign(q_d^T q) I_3 q_g^T q ,
            \end{align}
where $\sign$ denotes the signum function.

\todo{If the geodesic distance works better on all the examples, I vote for leaving out the error quadratic stuff.}

\section{Experiments} \label{sec:experiments}
    In this section we present several trajectory optimization problems for systems that
    undergo large changes in attitude: an airplane barrel roll, a quadrotor flip, and a
    satellite with flexible solar panels that must slew to a new orientation while avoiding a 
    keep-out zone. All results were run on a desktop computer with an 
    AMD Ryzen 2950x processor with 40 GB of RAM. 
    All problems are run using ALTRO, first without any of the modifications presented 
    in the current paper, analagous to the naive Newton's method in section \ref{sec:Wahbas} and labeled ``naive'', and then using the modifications listed
    in Sec. \ref{sec:trajopt} and the geodesic cost function described in Sec. \ref{sec:geodesic},
    labeled ``modified''.
    Timing results are summarized in Table \ref{tab:timing_results}.
    All experiments were solved to a constraint satisfaction tolerance of $10^{-5}$ and 
    discretized with a 4th order Runge-Kutta integrator.
    Code for all experiments is available on 
    \href{https://github.com/RoboticExplorationLab/PlanningWithAttitude}
    {GitHub\footnote{\url{https://github.com/RoboticExplorationLab/PlanningWithAttitude}}}.

    \begin{table}
        \centering
        \input{figures/timing_results.tex}
        \label{tab:timing_results}
        \caption{Trajectory Optimization Timing Results (naive/modified)}
    \end{table}
